---
title: "Challenge 1"
author: "310247934"
date: "2025-02-24"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(ggplot2)

```
```{r}
food_order <- read_csv("path")

```
## Initial Analysis
The following code provides an initial view of the mean preparation time of the meals on a weekday compared to a weekend. From the initial analysis there is not much difference when comparing the weekday against the weekend means. As such the use of the column graph whilst it does provide us with the global means does not provide much insight as where the variations occur.  

```{r}
food_order %>%
  group_by(`Day of the Week` = day_of_the_week) %>%
  summarise(`Number of Orders` = n(),
            `Average Prep Time` = mean(food_preparation_time)) %>%
  ggplot(aes(x = `Day of the Week`, y = `Average Prep Time`)) +
  geom_col(fill = "lightblue") +
  geom_text(aes(label = round(`Average Prep Time`, 1)), vjust = -0.5, size = 5) +
  labs(title = "Average Food Preparation Time by Day",
       x = "Day of the Week",
       y = "Average Prep Time") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  theme_minimal(base_size = 14)

```

## Further Insights
Through the use of faceting we are able to see the measure of spread of each cuisine type on a weekend compared to a weekday. This time of plot is more valuable as we can now see which cuisines have a wider spread of values and how the median value changes for each. The mean has also been overlaid represented by the blue dot. We can now see that the has a larger variance across the cuisine types. We can also see that the for many of the cuisines on both a weekday and weekend the food preparation time can vary any where from 20 to 35 minutes. The Thai and Vietnamese also had the largest difference between both their mean and median values when comparing the time taken on a weekday versus a weekend. 





```{r, fig.width=12, fig.height=8}
food_order %>%
  ggplot(aes(x = day_of_the_week, y = food_preparation_time, fill = day_of_the_week)) +
  geom_boxplot(show.legend = FALSE) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 4, color = "blue") +
  facet_wrap(~ cuisine_type) +
  labs(title = "Food Prep Time Distribution for Each Cuisine",
       x = "Day of the Week",
       y = "Food Preparation Time (Minutes)") +
  theme_minimal(base_size = 14)
```


